% arara: pdflatex
% !arara: biber
% !arara: pdflatex
% arara: pdflatex
% --------------------------------------------------------------------------
% the CNTFORMATS package
% 
%   A different way to read counters
% 
% --------------------------------------------------------------------------
% Clemens Niederberger
% Web:    https://bitbucket.org/cgnieder/exsheets/
% E-Mail: contact@mychemistry.eu
% --------------------------------------------------------------------------
% Copyright 2011-2014 Clemens Niederberger
% 
% This work may be distributed and/or modified under the
% conditions of the LaTeX Project Public License, either version 1.3
% of this license or (at your option) any later version.
% The latest version of this license is in
%   http://www.latex-project.org/lppl.txt
% and version 1.3 or later is part of all distributions of LaTeX
% version 2005/12/01 or later.
% 
% This work has the LPPL maintenance status `maintained'.
% 
% The Current Maintainer of this work is Clemens Niederberger.
% --------------------------------------------------------------------------
% The translations package is part of the exsheets bundle
% --------------------------------------------------------------------------
% If you have any ideas, questions, suggestions or bugs to report, please
% feel free to contact me.
% --------------------------------------------------------------------------
\documentclass[load-preamble+]{cnltx-doc}
\usepackage{cntformats}

\setcnltx{
  package  = {cntformats} ,
  authors  = Clemens Niederberger ,
  email    = {contact@mychemistry.eu} ,
  url      = {https://bitbucket.org/cgnieder/exsheets/} ,
  info     = {%
    {\small part of the \href{exsheets_en.pdf}{\ExSheets}
      bundle}\\[\baselineskip]
    a different way to read counters%
  } ,
  add-cmds = {
    @cntfmts@parsed@pattern, AddCounterPattern, 
    eSaveCounterPattern,  
    NewCounterPattern, NewPatternFormat,
    ReadCounterFrom, ReadCounterPattern, ReadCounterPatternFrom,
    SaveCounterPattern
  } ,
  add-silent-cmds = {
    alphalph,
    cs,
    ExplSyntaxOff, ExplSyntaxOn,
    int,
    myoddnumber,
    numexpr, ordinalnum,
    relax,
    somesetupcommand,
    tmpa, tmpb, the
  }
}

\defbibheading{bibliography}[\bibname]{\addsec{#1}}

\newpackagename\ExSheets{ExSheets}
\def\at{\cnltxat}

\usepackage[biblatex]{embrac}[2012/06/29]
  \ChangeEmph{[}[,.02em]{]}[.055em,-.08em]
  \ChangeEmph{(}[-.01em,.04em]{)}[.04em,-.05em]

\usepackage{booktabs,array,ragged2e}
\usepackage{subcaption}

% ----------------------------------------------------------------------------
% example definitions that have to be done in the preamble:
\usepackage{alphalph}
\NewPatternFormat{aa}{\alphalph}
\usepackage{fmtcount}
\NewPatternFormat{o}{\ordinalnum}
\newcommand*\myoddnumber[1]{\the\numexpr2*(#1)-1\relax}
\NewPatternFormat{x}{\myoddnumber}
\NewCounterPattern{testa}{ta}

\begin{document}

\section{Motivation}
\cntformats\ provides a way to format counters with what I will call patterns.
This does not in any way effect the usual \LaTeXe\ way of treating counters
and does not use \cs*{the\meta{counter}} nor is it affected by the redefinition of
them.

This package is aimed at package or class authors and probably not very useful
for document authors.

When I first had the idea for this package the idea of what it does already
existed as part of the \ExSheets\ package.  I can't recall why I came up with
the idea in the first place or why I originally wanted a new syntax for
formatting the \code{question} counter.  I am also not convinced any more that
it is a good idea.  Anyway, here we are.

\section{License and Requirements}\label{sec:license}
\license

\cntformats\ requires the \needpackage{etoolbox} package~\cite{pkg:etoolbox}
and the \pkg*{cnltx-base}\footnote{\CTANurl{cnltx}} package~\cite{bnd:cnltx}.

\section{Example}
A use case typically looks as follows:
\begin{example}[side-by-side]
  \ReadCounterPattern{se.sse}
\end{example}
where the key \code{se} stands for the current value of the \code{section}
counter and \code{sse} for \code{subsection}, respectively. \code{se.sse} is
an example for what will be called \emph{pattern}. The keys for the counters
can have optional arguments that specify the format:
\begin{example}[side-by-side]
  \stepcounter{subsection}
  \ReadCounterPattern{se[A](sse[r])}
\end{example}
\code{A} stands for \cs*{Alph} and \code{r} for \cs*{roman}.  A complete list
can be found in table~\ref{tab:predefined:formats} on
page~\pageref{tab:predefined:formats}.  As you can see you can insert
arbitrary other tokens in a pattern that won't be changed. It is important to
notice, though, that the patterns are only replaced if they're \emph{not}
placed in a braced group!

\begin{example}[side-by-side]
  \ReadCounterPattern{{se[A](sse[r])}}
\end{example}

I would imagine that the argument to \cs{ReadCounterPattern} is usually
supplied by a user setting an option \ldots
\begin{sourcecode}
  \somesetupcommand{
    counter-format = se[A](sse[r])
  }
\end{sourcecode}
\ldots{} and then internally used by the corresponding package or class.


\section{Usage}
In the following description of the available commands the symbol
\textcolor{expandable}{\expandablesign} means that the command is expandable.

In order to make counters known to \cntformats\ the following commands are used:
\begin{commands}
  \command{AddCounterPattern}[\sarg\oarg{module}\marg{counter}\marg{pattern}]
    This command will make the (existing) counter \meta{counter} known to
    \cntformats\ and assign the pattern \meta{pattern} to it.
  \command{NewCounterPattern}[\sarg\oarg{module}\marg{counter}\marg{pattern}]
    This command will create a new counter \meta{counter}, make it known to
    \cntformats\ and assign the pattern \meta{pattern} to it.
  \command{ReadCounterFrom}[\oarg{module}\marg{counter}\marg{internal cmd}]
    If you use one of the commands above with the starred version the number
    for the pattern is not automatically fetched from the internal
    \cs*{c@\meta{counter}}.  This can now be assigned with
    \cs{ReadCounterFrom}  where \meta{internal cmd} is the macro that holds
    the number.
\end{commands}
The commands above can only be used in the document preamble.

After the creation of these pattern markers one wants to be able to use them.
There are a number of macros that allow different aspects of usage.
\begin{commands}
  \command{ReadCounterPattern}[\oarg{module}\marg{pattern}]
    Reads, interprets and prints a pattern.
  \expandable\command{\at cntfmts\at parsed\at pattern}
    After \cs{ReadCounterPattern} has been used the current pattern
    interpretation is stored in this macro.  The \emph{interpretation} is
    \emph{not} what is printed.  See the examples below for details.
  \command{ReadCounterPatternFrom}[\oarg{module}\marg{macro that holds
    pattern}]
    Reads, interprets and prints a pattern that's stored in a macro.\\
    Otherwise the same as \cs{ReadCounterPattern}.
  \command{SaveCounterPattern}[\marg{cmd a}\marg{cmd b}\marg{pattern}]
    Saves the \meta{pattern} in \meta{cmd a} and the interpreted pattern in
    \meta{cmd b}.
  \command{eSaveCounterPattern}[\oarg{module}\marg{cmd a}\marg{cmd
    b}\marg{pattern}]
    Saves the \meta{pattern} in \meta{cmd a} and the expanded pattern in
    \meta{cmd b}.
  \command{SaveCounterPatternFrom}[\oarg{module}\marg{cmd a}\marg{cmd
    b}\marg{macro that holds pattern}]
    Like \cs{SaveCounterPattern} but reads the pattern from a macro.
  \command{eSaveCounterPatternFrom}[\oarg{module}\marg{cmd a}\marg{cmd
    b}\marg{macro that holds pattern}]
    Like \cs{eSaveCounterPattern} but reads the pattern from a macro.
\end{commands}

The optional argument \meta{module} should be specific for a package, say, so
that different patterns for the \code{section} for example don't interfer with
each other.  If you leave the argument the default module \code{cntfmts} is
used.

The \ExSheets\ packages uses the commands with the module \code{exsheets}.
You can find the following lines in \ExSheets' code:
\begin{sourcecode}
  \AddCounterPattern*[exsheets]{section}{se}
  \ReadCounterFrom[exsheets]{section} \l__exsheets_counter_sec_int
  \NewCounterPattern*[exsheets]{question}{qu}
  \ReadCounterFrom[exsheets]{question} \l__exsheets_counter_qu_int
\end{sourcecode}

Now let's see a short example that hopefully explains what the macros do:
\begin{example}
  % preamble
  % \NewCounterPattern{testa}{ta}
  \setcounter{testa}{11}
  \ReadCounterPattern{ta}
  \ReadCounterPattern{ta[a]} \\
  \ttfamily\makeatletter
  \meaning\@cntfmts@parsed@pattern
  
  \bigskip
  \SaveCounterPattern\tmpa\tmpb{ta[a]}
  \meaning\tmpa \\
  \meaning\tmpb
  
  \bigskip
  \eSaveCounterPattern\tmpa\tmpb{ta[a]}
  \meaning\tmpa \\
  \meaning\tmpb
\end{example}
You can see that somehow an additional \cs*{@empty} found its way into the
interpreted pattern. This is due to the fact that reading optional arguments
expandably isn't easy and must have some safety net.

\section{Predefined and New Patterns and Format Keys}
\subsection{Predefined Patterns and Format Keys}
\cntformats\ predefines a number of pattern keys. These are listed in
table~\ref{tab:predefined:patterns}.

\subsection{New Patterns and Format Keys}
Table~\ref{tab:predefined:formats} lists the predefined formats.  If you want
you can add own formats.
\begin{commands}
  \command{NewPatternFormat}[\marg{pattern}\marg{format}]
    \meta{format} is a number presentation command like \cs*{@alph}, \ie,
    it needs a mandatory argument that takes a number.  It is used in
    \meta{format} \emph{without} its argument.  This command can only be used
    in the preamble.
\end{commands}

\begin{table}
  \centering
  \caption{Predefined Patterns and Format Keys.}
  \addtocounter{table}{-1}% necessary fix for subcaption and KOMA
  \begin{minipage}{.48\linewidth}
    \centering
    \subcaption{Predefined Patterns for the module \code{cntfmts}.}
    \label{tab:predefined:patterns}
    \begin{tabular}{>{\ttfamily}l>{\ttfamily}l}
      \toprule
        \normalfont\bfseries counter & \normalfont\bfseries pattern \\
      \midrule
        chapter       & ch \\
        section       & se \\
        subsection    & sse \\
        subsubsection & ssse \\
        paragraph     & pg \\
     \bottomrule
    \end{tabular}
  \end{minipage}
  \begin{minipage}{.48\linewidth}
    \centering
    \subcaption{Predefined Format Keys}
    \label{tab:predefined:formats}
    \begin{tabular}{>{\ttfamily}ll}
      \toprule
        \normalfont\bfseries key & \normalfont\bfseries format \\
      \midrule
        1 & \cs*{arabic} \\
        a & \cs*{alph} \\
        A & \cs*{Alph} \\
        r & \cs*{roman} \\
        R & \cs*{Roman} \\
     \bottomrule
    \end{tabular}
  \end{minipage}
\end{table}

Here are now a few examples of possible new patterns. Suppose the following code
in the preamble:
\begin{sourcecode}
  \usepackage{alphalph,fmtcount}
  \newcommand*\myoddnumber[1]{\the\numexpr2*(#1)-1\relax}

  \NewPatternFormat{aa}{\alphalph}
  \NewPatternFormat{o}{\ordinalnum}
  \NewPatternFormat{x}{\myoddnumber}

  \newcounter{testa}
  \NewCounterPattern{testa}{ta}
  \setcounter{testa}{4}
\end{sourcecode}

Then we can use the new pattern and the new formats as
follows:\setcounter{testa}{4}
\begin{example}
  \ReadCounterPattern{ta[aa]}
  \ReadCounterPattern{ta[o]}
  \ReadCounterPattern{ta[x]}
\end{example}

\end{document}
